A formal treatment of bidirectional typing

### 2023-10-16 tex/ESOP24/BiSig.tex
[ tex ] Update Future work
[ tex ] Update after sync in person

\begin{abstract}
There has been much progress in designing bidirectional type systems and associated type synthesis algorithms, but mainly on a case-by-case basis.
This situation is in stark contrast to parsing, for which there have been general theories and widely applicable tools such as parser generators.
To remedy the situation, this paper develops a \emph{general} and \emph{formal} theory of bidirectional typing, and provides a \emph{verified} generator of \emph{proof-relevant} type synthesisers for simply typed languages: For every signature that specifies a mode-correct bidirectionally typed language, there exists a proof-relevant type synthesiser that for an input abstract syntax tree constructs a typing derivation if any, gives its refutation if not, or reports that the input does not have enough type annotations.
Soundness, completeness, and mode-correctness are studied universally for any simply typed language, which are sufficient conditions for deriving a type synthesiser.
We propose a preprocessing step called \emph{mode decoration}, which helps the user to deal with missing type annotations in a given abstract syntax tree.
The entire development is formalised in \Agda and can be further integrated with other language-formalisation frameworks.
\end{abstract}

\input{S1-intro}
\input{S2-outline}
\input{S3-defs.tex}
\input{S4-pre-synthesis}
\input{S5-synthesis}
\input{S6-example}
\subfile{S7-formalisation}
\input{S8-future}

@@

\appendix
\input{appendix}

>>>>

\begin{abstract}
There has been much progress in designing bidirectional type systems and associated type synthesis algorithms, but mainly on a case-by-case basis.
%This situation is in stark contrast to parsing, for which there have been general theories and widely applicable tools such as parser generators.
To remedy the situation, this paper develops a \emph{general} and \emph{formal} theory of bidirectional typing and as a by-product of our formalism provides a \emph{verified} generator of \emph{proof-relevant} type synthesisers for simply typed languages: For every signature that specifies a mode-correct bidirectionally typed language, there exists a proof-relevant type synthesiser that for an input abstract syntax tree constructs a typing derivation if any, gives its refutation if not, or reports that the input does not have enough type annotations.
Soundness, completeness, and mode-correctness are studied universally for al signatures, which are sufficient conditions for deriving a type synthesiser.
We propose a preprocessing step called \emph{mode decoration}, which helps the user to deal with missing type annotations in a given abstract syntax tree.
The entire development is formalised in \Agda and can be further integrated with other language-formalisation frameworks.
\end{abstract}

\input{S1-intro}
\input{S2-outline}
\input{S3-defs.tex}
\input{S4-pre-synthesis}
\input{S5-synthesis}
\input{S6-example}
\input{S8-future}

@@

\appendix

%\input{appendix}
\subfile{S7-formalisation}

### 2023-10-16 tex/ESOP24/S1-intro.tex
[ tex ] Update after sync in person

Our work adopts a proof-relevant approach to (bidirectional) type synthesis, as illustrated by \citet{Wadler2022} for \PCF.
The proof-relevant formulation deviates from the usual one:
traditionally, a type synthesis algorithm is presented as \emph{algorithmic rules}, for example in the form $\Gamma |- \isTerm{t} \syn A \mapsto \isTerm{t}'$, which denotes that annotations can be added to~$\isTerm{t}$ in the surface language%
\Josh{The algorithm adds annotations to the input term?}
to produce $\isTerm{t}'$ of type~$A$ in the core language.
Such an algorithm is then accompanied by soundness and completeness assertions that the algorithm correctly synthesises the type of a raw term and every typable term can be synthesised.
By contrast, the proof-relevant approach exploits the simultaneously computational and logical nature of Martin-L\"of type theory and formulates \emph{algorithmic soundness, completeness, and decidability in one go}.
We explain this in a bit more detail below.
%(although they do not emphasise the difference of their approach from the traditional one)

Recall that the law of excluded middle $P + \neg P$ does not hold as an axiom for every~$P$ constructively, and we say that $P$~is logically \emph{decidable} if the law holds for~$P$.
Since Martin-L\"of type theory is simultaneously logical and computational, a decidability proof is also a proof-relevant decision procedure that computes not only a yes-or-no answer but also a proof of~$P$ or its refutation, so logical decidability is also algorithmic decidability.
%For example, a proof of \emph{decidable equality}, i.e.~$\forall x, y.~(x = y) + (x \neq y)$, decides whether $x$~and~$y$ are equal and accordingly gives an identity proof or a refutation explicitly.
%Such a decidability proof may or may not be possible depending on the domain of $x$~and~$y$, and is non-trivial in general.
%In type theory all proofs as programs terminate, so logical decidability implies algorithmic decidability.
More specifically, consider the statement of the type inference problem
\begin{quote}
  `for a context $\Gamma$ and a raw term $t$, either a typing derivation of\, $\Gamma |- t : A$ exists for some type~$A$ or any derivation of\, $\Gamma |- t : A$ for some type~$A$ leads to a contradiction',
\end{quote}
which can be rephrased more succinctly as
\begin{quote}
  `It is \underline{\emph{decidable}} for any $\Gamma$~and~$t$ whether $\Gamma |- t : A$ is derivable for some~$A$'.
\end{quote}
A proof of this statement would also be a program that produces either a typing derivation for the given raw term~$\isTerm{t}$ or a negation proof that such a derivation is impossible.
The first case is algorithmic soundness, while the second case is algorithmic completeness in contrapositive form (which implies the original form due to the decidability).
Hence, proving the statement is the same as constructing a verified proof-relevant type inference algorithm, which returns not only an answer but also a proof justifying that answer.
This is an economic way to bridge the gap between theory and practice, where proofs double as verified programs, in contrast to separately exhibiting a theory and an implementation that are only loosely related.

@@

\subsubsection{Mode-correctness and general definitions of languages}
\label{sec:language-formalisation}
The most essential characteristics of bidirectional typing is \emph{mode-correctness}, since an algorithm can often be `read off' from the definition of a bidirectionally typed language if mode-correct.
It seems that the implications of mode-correctness have only been addressed informally by \citet{Dunfield2021}, since mode-correctness is not formally defined as a \emph{property of languages}.
%But, what is a language anyway?

To make the notion of mode-correctness precise, we first give a general definition of bidirectional simple type systems, called \emph{bidirectional binding signature}, extending the sorted version of \varcitet{Aczel1978}{'s binding signature} with bidirectionality.
A general definition of typed languages allows us to define mode-correctness and to investigate its consequences rigorously, including the uniqueness of synthesised types and the decidability of bidirectional type synthesis for all mode-correct signatures.
The proof of the latter theorem amounts to a generator of proof-relevant bidirectional type synthesisers (analogous to a parser generator working for unambiguous or disambiguated grammars).

To make our exposition accessible, the theory in this paper focuses on simply typed languages with a syntax-directed bidirectional type system, so that the decidability of bidirectional type synthesis can be established without any other technical assumptions.
%the idea of extending a signature with bidirectionality should be clear enough for further generalisation.
By adding assumptions, it should be possible to extend the theory to deal with more expressive type systems; for instance, we briefly discuss how the theory can handle polymorphically typed languages such as \SystemF and \SystemFsub with additional assumptions in \cref{sec:future}.

@@

%
In short, we develop a general and formal theory of bidirectional type synthesis for simply typed languages, including 
\begin{enumerate}
  \item general definitions for bidirectional type systems and mode-correctness;
  \item mode derivations for explicitly dealing with annotations in the theory, and mode decoration for helping the user to work with annotations in practice;
  \item rigorously proven consequences of mode-correctness---the uniqueness of synthesised types and the decidability of bidirectional type synthesis, which amounts to
  \item a fully verified generator of proof-relevant type-synthesisers.
\end{enumerate}
Our theory is fully formally developed in \Agda but translated to the mathematical vernacular for presentation in this paper.
The formal theory doubles as a verified implementation, with no gap in between.

@@

This paper is structured as follows.
We first present a concrete overview of our theory using simply typed $\lambda$-calculus in \cref{sec:key-ideas}, prior to developing a general framework for specifying bidirectional type systems in \cref{sec:defs}.
Following this, we discuss mode decoration and related properties in \cref{sec:pre-synthesis}.
The main technical contribution lies in \cref{sec:type-synthesis}, where we introduce mode-correctness and bidirectional type synthesis.
Some examples other than simply typed $\lambda$-calculus are given in \Cref{sec:example}.
We briefly demonstrate the use of our \Agda formalisation as programs in \cref{sec:formalisation} and conclude in \cref{sec:future} with further developments.

>>>>

Our work adopts a proof-relevant approach to (bidirectional) type synthesis, as illustrated by \citet{Wadler2022} for \PCF.
The proof-relevant formulation deviates from the usual one:
traditionally, a type synthesis algorithm is presented as \emph{algorithmic rules}, for example in the form $\Gamma |- \isTerm{t} \syn A \mapsto \isTerm{t}'$, which denotes that $\isTerm{t}$ in the surface language
%\Josh{The algorithm adds annotations to the input term?}
can be transformed to a well typed term $\isTerm{t}'$ of type~$A$ in the core language~\cite{Pierce2000}.
Such an algorithm is then accompanied by soundness and completeness assertions that the algorithm correctly synthesises the type of a raw term and every typable term can be synthesised.
By contrast, the proof-relevant approach exploits the simultaneously computational and logical nature of Martin-L\"of type theory and formulates \emph{algorithmic soundness, completeness, and decidability in one go}.
%(although they do not emphasise the difference of their approach from the traditional one)

In more detail:
Recall that the law of excluded middle $P + \neg P$ does not hold as an axiom for every~$P$ constructively, and we say that $P$~is logically \emph{decidable} if the law holds for~$P$.
Since Martin-L\"of type theory is simultaneously logical and computational, a decidability proof is also a proof-relevant decision procedure that computes not only a yes-or-no answer but also a proof of~$P$ or its refutation, so logical decidability is also algorithmic decidability.
%For example, a proof of \emph{decidable equality}, i.e.~$\forall x, y.~(x = y) + (x \neq y)$, decides whether $x$~and~$y$ are equal and accordingly gives an identity proof or a refutation explicitly.
%Such a decidability proof may or may not be possible depending on the domain of $x$~and~$y$, and is non-trivial in general.
%In type theory all proofs as programs terminate, so logical decidability implies algorithmic decidability.
More specifically, consider the statement of the type inference problem
\begin{quote}
  `for a context $\Gamma$ and a raw term $t$, either a typing derivation of\, $\Gamma |- t : A$ exists for some type~$A$ or any derivation of\, $\Gamma |- t : A$ for some type~$A$ leads to a contradiction',
\end{quote}
which can be rephrased more succinctly as
\begin{quote}
  `It is \underline{\emph{decidable}} for any $\Gamma$~and~$t$ whether $\Gamma |- t : A$ is derivable for some~$A$'.
\end{quote}
A proof of this statement would also be a program that produces either a typing derivation for the given raw term~$\isTerm{t}$ or a negation proof that such a derivation is impossible.
The first case is algorithmic soundness, while the second case is algorithmic completeness in contrapositive form (which implies the original form due to the decidability).
Hence, proving the statement is the same as constructing a verified proof-relevant type inference algorithm, which returns not only an answer but also a proof justifying that answer.
This is an economic way to bridge the gap between theory and practice, where proofs double as verified programs, in contrast to separately exhibiting a theory and an implementation that are only loosely related.

@@

\subsubsection{Mode-correctness and general definitions of languages}
\label{sec:language-formalisation}
The most essential characteristics of bidirectional typing is \emph{mode-correctness}, since an algorithm can often be `read off' from the definition of a bidirectionally typed language if mode-correct.
As illustrated by \citet{Dunfield2021}, it seems that the implications of mode-correctness have only been addressed informally so far and mode-correctness is not yet formally defined as a \emph{property of languages}.

In order to make the notion of mode-correctness precise, we first give a general definition of bidirectional simple type systems, called \emph{bidirectional binding signature}, extending the sorted version of \varcitet{Aczel1978}{'s binding signature} with modes.
A general definition of typed languages allows us to define mode-correctness and to investigate its consequences rigorously, including the uniqueness of synthesised types and the decidability of bidirectional type synthesis for all mode-correct signatures.
The proof of the latter theorem amounts to a generator of proof-relevant bidirectional type synthesisers (analogous to a parser generator working for unambiguous or disambiguated grammars).

To make our exposition accessible, the theory in this paper focuses on simply typed languages with a syntax-directed bidirectional type system, so that the decidability of bidirectional type synthesis can be established without any other technical assumptions.
%the idea of extending a signature with bidirectionality should be clear enough for further generalisation.
It should be possible to extend the theory to deal with more expressive types and assumptions other than mode-correctness.
For instance, we briefly discuss how the theory can handle polymorphically typed languages such as \SystemF, \SystemFsub, and those systems using implicit type applications with additional assumptions in \cref{sec:future}.

@@

%
In short, we develop a general and formal theory of bidirectional type synthesis for simply typed languages, including 
\begin{enumerate}
  \item general definitions for bidirectional type systems and mode-correctness;
  \item mode derivations for explicitly dealing with annotations in the theory, and mode decoration for helping the user to work with annotations in practice;
  \item rigorously proven consequences of mode-correctness---the uniqueness of synthesised types and the decidability of bidirectional type synthesis, which amounts to
  \item a fully verified generator of proof-relevant type-synthesisers.
\end{enumerate}
Our theory is fully formally developed with \Agda but translated to the mathematical vernacular for presentation in this paper.
The formal theory doubles as a verified implementation, with no gap in between.

@@

This paper is structured as follows.
We first present a concrete overview of our theory using simply typed $\lambda$-calculus in \cref{sec:key-ideas}, prior to developing a general framework for specifying bidirectional type systems in \cref{sec:defs}.
Following this, we discuss mode decoration and related properties in \cref{sec:pre-synthesis}.
The main technical contribution lies in \cref{sec:type-synthesis}, where we introduce mode-correctness and bidirectional type synthesis.
Some examples other than simply typed $\lambda$-calculus are given in \Cref{sec:example}.
We conclude in \cref{sec:future} with further developments and a demonstration in \cref{sec:formalisation} for the use of our \Agda formalisation as programs.

### 2023-10-16 tex/ESOP24/S8-future.tex
[ tex ] Update after sync in person

\section{Discussion} \label{sec:future}
We believe that our formal treatment lays a foundation for further investigation, as the essential aspects of bidirectional typing have been studied rigorously. 
While our current development is based on simply typed languages to highlight the core ideas, it is evident that many concepts and extensions remain unexplored.
In this section we discuss related work and outline potential directions for future research.
\subsubsection{Language formalisation frameworks}
Many general definitions of logics and type theories and frameworks for defining them have been proposed.

@@

%\subsubsection{Generic bidirectional type synthesis beyond simple types}
%Clearly, bidirectional type synthesis becomes prominent because it can scale up to deal with more complex types than simple types.
%As for a general treatment for another class of languages, one has to devise a general definition of languages, which may have been used by existing language-formalisation frameworks.
%Then, one could enrich the general definition with modes and adopt the definition of mode-correctness accordingly. 
%Soundness and completeness of our formulation should carry over, since they are only about the separation and combination of mode and typing information for a given raw term (in a syntax-directed formulation).
%Mode decoration is about annotating a raw term with modes and marking missing annotations, used for guiding the direction during bidirectional type synthesis, so it should also carry over.
%What makes the general development of bidirectional type synthesis challenging is inherent to the tension between expressiveness and usability for type synthesis and the general treatment for other language facilities.
%We discusses the cases for dependent types and for polymorphic types below.
%
%\paragraph{Dependent types} Type synthesis for a dependent type theory involves term equality check for type equality problem, so a generic decidable bidirectional type synthesis relies on the decidability of the term equality check which in turn requires the ability of specifying computation rules within a language specification.
%\varcitet{Felicissimo2023}{'s computational LF} is capable of defining dependent type theories with computational rules like \textsc{Dedukti}~\cite{Assaf2016} and generic bidirectional type synthesis.
%Apart from their more general definition targeting dependent type theories, generic bidirectional type synthesis \cite[Theorem~3.12]{Felicissimo2023} that is decidable, sound, and complete (with respect to decorated terms) is possible under the assumption that a given language specification is mode-correct (or well-typed \cite[Section~3.2]{Felicissimo2023}) and equality check is well-behaved: the specified computation rules are type-preserving, confluent, and strongly normalising.
%Yet, the last assumption is hard to establish even for a specific language and still lacks a general consideration.
%
%\paragraph{Polymorphic types} For \SystemF and other languages where type-level variable binding is allowed, we can start with the notion of polymorphic signature $(\Sigma, \Omega)$ by~\citet{Hamana2011} whereas the arity of each type construct $\sigma$ in $\Sigma$ becomes a binding arity with a single sort $*$ and accordingly a term construct can specify a pair of extension contexts~$\left<\Xi\right>[\Delta]$ for term variables $\Delta$ and type variables $\Xi$.
%It is not hard to see how general definitions for bidirectional typing rules and mode derivations can be extended from \varcitet[Section~3.4]{Hamana2011}{'s definitions}.
%For example, the universal type $\forall \alpha.\, A$ and type abstraction in \SystemF can then be specified as operations
%\[
%  \mathsf{all}\colon *\rhd [*]* \to *
%  \quad\text{and}\quad
%  \mathsf{tabs}\colon [*]A \rhd \left< * \right>A^{\chk} \to
%  \tyOp_{\mathsf{all}}(\alpha.\, A)^{\chk}
%\]
%\Cref{thm:bidirectional-type-synthesis-checking} should also carry over, since no equation is imposed on types and no guess work is needed.
%However, explicit type application is unusable in practice but implicit application results the \emph{instantiation problem}~\cite[Section~5]{Dunfield2021} which amounts to replacing type equality check with checking the subtyping relation $A <: B$ in rule $\ChkRule{Sub}$ and type synthesis with the guess of $B$ in $\forall \alpha.\, A <: A[B/\alpha]$. 
%A satisfactory theory for bidirectional typing with polymorphism and subtyping should be neutral and thus accommodate various solutions to the instantiation problem, which is still under development.
%
\subsubsection{Generic bidirectional type synthesis beyond simple types}

Bidirectional type synthesis plays a crucial role in handling more complex types than simple types.
To provide a formal treatment for a broader class of languages, we need a general definition of languages in question, which may have been used in existing theories or frameworks.
Then, this general definition can be augmented with modes and the concept of mode-correctness can be introduced.
Soundness and completeness (\cref{lem:soundness-completeness}) should naturally extend, as they pertain solely to the separation and combination of mode and typing information within a given raw term in a syntax-directed formulation.
Mode decoration (\cref{sec:mode-decoration}) involves annotating a raw term with modes and marking missing annotations should also work.
The challenge in developing bidirectional type synthesis on a more general case comes from striking the right balance between expressiveness and usability for type synthesis while considering other language features.
We discuss cases involving dependent types and polymorphic types below.

\paragraph{Dependent Types}
In the context of dependent type theory, type synthesis entails a term equality check for type equality.
Achieving generic decidable bidirectional type synthesis depends on the decidability of term equality, which, in turn, relies on the ability to specify computation rules within a language specification.

\varcitet{Felicissimo2023}{'s computation LF} is capable of defining dependent type theories with computational rules, similar to \Dedukti, and enabling generic bidirectional type synthesis.
Generic bidirectional type synthesis, as presented in Felicissimo's work, is decidable, sound, and complete with respect to decorated terms, provided that the underlying language specification is mode-correct, and the equality check is well-behaved, i.e.\ type-preserving, confluent, and strongly normalising.
However, the last assumption is challenging to establish, even for specific languages and lacks a comprehensive general understanding.

\paragraph{Polymorphic Types}
In the case of languages like \SystemF and others that permit type-level variable binding, we can start with the notion of a polymorphic signature, as introduced by Hamana---each type construct in a signature is specified by a binding arity with a single sort $*$ and a term construct can employ a pair of extension contexts for term variables and type variables.

Extending general definitions for bidirectional typing and mode derivations from \citeauthor{Hamana2011}'s work is straightforward. 
For example, the universal type $\forall \alpha.\, A$ and type abstraction in \SystemF can be specified as operations:
\[
\mathsf{all} : * \rhd [*] * \to *
\quad\text{and}\quad
\mathsf{tabs} : [*] A \rhd \left< * \right> A^{\chk} \to \tyOp_{\mathsf{all}}(\alpha. A)^{\chk}
\]
The decidability of bidirectional type synthesis (\cref{thm:bidirectional-type-synthesis-checking}) should also carry over, as no equations are imposed on types and no guessing (for type application) is required.
Adding subtyping $A \mathrel{<:} B$ to languages can be done by replacing type equality with a subtyping relation $\mathrel{<:}$ and type equality check with subtyping check, so polymorphically typed languages with subtyping such as \SystemFsub can be specified.
The main idea of bidirectional typing does not change, so the formal development should extend without further assumptions as well.

However, explicit type application is impractical but its implicit counterpart results in the \emph{instantiation problem} which amounts to transforming the implicit type application (a stationary rule) to guessing $B$ in $\forall \alpha. A <: A[B/\alpha]$.
A general theory for bidirectional typing that accommodates various solutions to the instantiation problem as an assumption is still under development.

@@

\subsubsection{Conclusion}
There are still many concepts that are possible to articulate such as, for example, Pfenning's recipe for bidirectionalising typing rules.
In contrast, there are concepts that may be hard to pin down, notably `annotation character'~\cite{Dunfield2021}, which is roughly about how easy it is for the user to write annotated programs.
These considerations may continue to require creativity from the designer of type systems whereas its implementation could be automated entirely.

>>>>

\section{Discussion} \label{sec:future}
We believe that our formal treatment lays a foundation for further investigation, as the essential aspects of bidirectional typing have been studied rigorously. 
While our current development is based on simply typed languages to highlight the core ideas, it is evident that many concepts and aspects remain unexplored.
In this section we discuss related work and potential directions for future research.
\subsubsection{Language formalisation frameworks}
Many general definitions of logics and type theories and frameworks for defining them have been proposed.

@@

\subsubsection{General theory of bidirectional type synthesis beyond simple types}

Bidirectional type synthesis plays a crucial role in handling more complex types than simple types, and we sketch how our theory can be extended to treat a broader class of languages.
We need a general definition of languages in question (\cref{subsec:simple-types,subsec:binding-sig}).
Then, this general definition can be augmented with modes (\cref{subsec:bidirectional-system}) and the formal definition of mode-correctness can be adapted (\cref{def:mode-correctness}).
Soundness and completeness (\cref{lem:soundness-completeness}) should still hold, as they amount to the separation and combination of mode and typing information for a given raw term (in a syntax-directed formulation).
Mode decoration (\cref{sec:mode-decoration}) involves annotating a raw term with modes and marking missing annotations, and should also work.
As for the decidability of bidirectional type synthesis, we discuss two cases involving polymorphic types and dependent types below.

\paragraph{Polymorphic types}
In the case of languages like \SystemF and others that permit type-level variable binding, we can start with the notion of a polymorphic signature, as introduced by \citet{Hamana2011}:
\begin{inlineenum}
  \item each type construct in a signature is specified by a binding arity with a single sort $*$, and
  \item a term construct can employ a pair of extension contexts for term variables and type variables.
\end{inlineenum}

Extending general definitions for bidirectional typing and mode derivations from \citeauthor{Hamana2011}'s work is straightforward. 
For example, the universal type $\forall \alpha.\, A$ and type abstraction in \SystemF can be specified as operations:
\[
\mathsf{all} : * \rhd [*] * \to *
\quad\text{and}\quad
\mathsf{tabs} : [*] A \rhd \left< * \right> A^{\chk} \to \tyOp_{\mathsf{all}}(\alpha. A)^{\chk}
\]
The decidability of bidirectional type synthesis (\cref{thm:bidirectional-type-synthesis-checking}) should also carry over, as no equations are imposed on types and no guessing (for type application) is required.
Adding subtyping $A \mathrel{<:} B$ to languages can be done by replacing type equality with a subtyping relation $\mathrel{<:}$ and type equality check with subtyping check, so polymorphically typed languages with subtyping such as \SystemFsub can be specified.
The main idea of bidirectional typing does not change, so it should be possible to extend the formal theory without further assumptions too.

However, explicit type application is impractical but its implicit counterpart results in the \emph{instantiation problem} which amounts to transforming the implicit type application (a stationary rule) to guessing $B$ in $\forall \alpha. A <: A[B/\alpha]$.
A general theory for bidirectional typing that accommodates various solutions to the instantiation problem as an assumption is left as future work.

\paragraph{Dependent types}
In the context of dependent type theory, type synthesis entails a term equality check for type equality.
Achieving decidable bidirectional type synthesis depends on the decidability of term equality, which, in turn, relies on the ability to specify computation rules within a language specification.

Extending the logical framework \Dedukti, \varcitet{Felicissimo2023}{'s \textsf{CompLF}} is capable of defining dependent type theories with computational rules and enables generic bidirectional type synthesis.
\LT{compare his work with ours}
Generic bidirectional type synthesis, as presented in Felicissimo's work, is decidable, sound, and complete with respect to mode-decorated terms, provided that the underlying language specification is mode-correct, and the equality check is \emph{well-behaved}, i.e.\ type-preserving, confluent, and strongly normalising.
However, the last assumption is challenging to establish even for specific languages and lacks a comprehensive general understanding.

@@

There are still many concepts that are possible to articulate such as, for example, Pfenning's recipe for bidirectionalising typing rules.

In contrast, there are concepts that may be hard to pin down, notably `annotation character'~\cite{Dunfield2021}, which is roughly about how easy it is for the user to write annotated programs.
These considerations may continue to require creativity from the designer of type systems whereas its implementation could be automated entirely.

\begin{table}[ht]
  \renewcommand{\arraystretch}{2.5}
  \setlength{\tabcolsep}{3pt}
  \centering\footnotesize
\begin{tabular}{c | l}
  Rules & Operations \\ \hline\hline
  $\inferrule{ }{\Gamma \vdash \mathtt{z} \chk \mathtt{nat}}$ &  $\aritysymbol{\mathsf{z}}{\cdot}{\cdot}{\mathtt{nat}^{\chk}}$ 
  \\
 $\inferrule{\Gamma \vdash t \chk \mathtt{nat}}{\Gamma \vdash \mathtt{s}(t) \chk \mathtt{nat}}$ & $\aritysymbol{\mathsf{s}}{\cdot}{\mathtt{nat}^{\chk}}{\mathtt{nat}^{\chk}}$ \\
 $\inferrule{\Gamma \vdash t \syn \mathtt{nat} \\\\ \Gamma \vdash t_0 \chk A \and \Gamma, x : \mathtt{nat} \vdash t_1 \chk A}{\Gamma \vdash \mathtt{ifz}(t_0; x.t_1)(t) \chk A}$ & $\aritysymbol{\mathsf{ifz}}{A}{\mathtt{nat}^{\syn}, A^{\chk}, A^{\chk}}{A^{\chk}}$ \\
 $\inferrule{\Gamma \vdash t \chk A \and \Gamma \vdash u \chk B}{\Gamma \vdash (t, u) \chk A \times B}$ & $\aritysymbol{\mathsf{pair}}{A, B}{A^{\chk}, B^{\chk}}{A \times B^{\chk}}$  \\
 $\inferrule{\Gamma \vdash t \syn A_1 \times A_2}{\Gamma \vdash \mathtt{proj}_i(t) \syn A_i}$ for $i = 1, 2$ & $\aritysymbol{\mathsf{proj}_i}{A_1, A_2}{A_1 \times A_2^{\syn}}{A_i^{\syn}}$ for $i = 1, 2$ \\

 $\inferrule{\Gamma \vdash t \chk A_i}{\Gamma \vdash \mathtt{inj}_i(t) \chk A_1 + A_2}$ for $i = 1, 2$ & $\aritysymbol{\mathsf{inj}_i}{A_1, A_2}{A_i^{\chk}}{A_1 + A_2^{\chk}}$ for $i = 1, 2$ \\

 $\inferrule{\Gamma \vdash u \syn A + B \\\\ \Gamma, x_1 : A \vdash t_1 \chk C \and \Gamma, x_2 : B \vdash t_2 \chk C}{\Gamma \vdash \mathtt{case}(u; x_1. t_1; x_2. t_2) \chk C}$ & $\aritysymbol{\mathsf{case}}{A, B, C}{{A + B}^{\syn}, [A]C^{\chk}, [B]C^{\chk}}{C^{\syn}}$ \\

 $\inferrule{\Gamma, x : A \vdash t \chk A}{\Gamma \vdash \mu x.\, t \chk A}$ & $\aritysymbol{\mathsf{mu}}{A}{[A]A^{\chk}}{A^{\chk}}$ \\

 $\inferrule{\Gamma \vdash t \syn A \and \Gamma, x : A \vdash u \chk B}{\Gamma \vdash \mathtt{let}\;x = t\;\mathtt{in}\;u\chk B}$ & 
  $\aritysymbol{\mathsf{let}}{A, B}{A^{\syn},[A]B^{\chk}}{B^{\chk}}$ \\
 $\inferrule{\Gamma \vdash t \syn A}{\Gamma \vdash \mathtt{ret}(t) \syn T(A)}$ & $\aritysymbol{\mathsf{ret}}{A}{A^{\syn}}{T(A)^{\syn}}$ \\
 $\inferrule{\Gamma \vdash t \syn T(A) \and \Gamma, x : A \vdash u \syn T(B)}{\Gamma \vdash \mathtt{bind}(t; x.u) \syn T(B)}$ & $\aritysymbol{\mathsf{bind}}{A, B}{T(A)^{\syn}, [A]T(B)^{\syn}}{T(B)^{\syn}}$
\end{tabular}
\caption{A computational calculus with naturals, products, sums, and general recursion}
\label{tab:computational-calculus}
\end{table}
